Combinatorial Proofs and Decomposition Theorems for First-order Logic

### 2021-01-22 main.tex
rule permutation

  We have the following derivations:
  $$
  \vlderivation{
    \vlin{\cD}{}{(A \vlor B) \vlan (C \vlor D)}{
      \vlin{\fequ}{}{((A \vlor B) \vlan (C \vlor D)) \vlor ((A \vlor B) \vlan (C \vlor D))}{
      \vlin{\wrD}{}{((A \vlor B) \vlan (C \vlor D)) \vlor ((B \vlor A) \vlan (D \vlor C))}{
	\vlin{\wrD}{}{((A \vlor B) \vlan (C \vlor D)) \vlor ((B \vlor A) \vlan D))}{
	  \vlin{\wrD}{}{((A \vlor B) \vlan (C \vlor D)) \vlor (B \vlan D))}{
	    \vlin{\wrD}{}{((A \vlor B) \vlan C) \vlor (B \vlan D)}{
	      \vlhy{(A \vlan C) \vlor (B \vlan D)}}}}}}}}
  $$.
  $$
  \vlderivation{
    \vlin{\cD}{}{\exists x.(A \vlor B)}{
      \vlin{\fequ}{}{\exists x.(A \vlor B)) \vlor (\exists x.(A \vlor B))}{
      \vlin{\wrD}{}{(\exists x.(A \vlor B)) \vlor (\exists x.(B \vlor A))}{
	\vlin{\wrD}{}{(\exists x.(A \vlor B)) \vlor (\exists x.B)}{
	  \vlhy{(\exists x.A) \vlor (\exists x.B)}}}}}}
  $$.
  $$
  \vlderivation{
    \vlin{\cD}{}{\forall x.(A \vlor B)}{
      \vlin{\fequ}{}{(\forall x.(A \vlor B)) \vlor (\forall x.(A \vlor B))}{
      \vlin{\wrD}{}{(\forall x.(A \vlor B)) \vlor (\forall x.(B \vlor A))}{
	\vlin{\wrD}{}{(\forall x.(A \vlor B)) \vlor (\forall x.B)}{
	  \vlhy{(\forall x.A) \vlor (\forall x.B)}}}}}}
  $$.
  \juihsuan{done. If needed, we can introduce the notion of open deduction to
reduce the size of derivations...}\lutz{I was thinking about that, but (i) it is probably not worth the effort, as we wont have many derivation, and (ii) it is hard to define rectified derivations this way.}
\end{proof}

@@

  Suppose that $\Gamma$ is provable in $\FOLK$, then so is $\form{\Gamma}$.
Consider a proof
$ \vlderivation{
        \vltrl{}{\FOLK}{\Deri}{\sqns{\form{\Gamma}}}{
          \vlhy{\quad}}{
          \vlhy{}}{
          \vlhy{}}} $

@@

\begin{equation*}
\vlderivation{
  \vliin{\vlan}{}{\sqns{\Gamma, A \vlan B, \Delta}}{
    \vlin{\weakr}{}{\sqns{\Gamma, A}}{
      \vlhy{\sqns{\Gamma}}}}{
    \vlhy{\sqns{B, \Delta}}}}
\leadsto
\vlderivation{
  \vlin{\weakr}{}{\sqns{\Gamma, A \vlan B, \Delta}}{
    \vlin{\weakr}{}{\vlvdots}{
      \vlhy{\sqns{\Gamma}}}}}
\end{equation*}
\begin{equation*}
\vlderivation{
  \vlin{\vlor}{}{\sqns{\Gamma, A \vlor B}}{
    \vlin{\weakr}{}{\sqns{\Gamma, A, B}}{
      \vlhy{\sqns{\Gamma, A}}}}}
\leadsto
\vlderivation{
  \vlin{\wrD}{}{\sqns{\Gamma, A \vlor B}}{
    \vlhy{\sqns{\Gamma, A}}}}
\end{equation*}
\begin{equation*}
\vlderivation{
  \vlin{\exists}{}{\sqns{\Gamma, \exists x.A}}{
    \vlin{\weakr}{}{\sqns{\Gamma, A\ssubst{x}{t}}}{
      \vlhy{\sqns{\Gamma}}}}}
\leadsto
\vlderivation{
  \vlin{\weakr}{}{\sqns{\Gamma, A\ssubst{x}{t}}}{
    \vlhy{\sqns{\Gamma}}}}
\end{equation*}
\begin{equation*}
\vlderivation{
  \vlin{\forall}{}{\sqns{\Gamma, \forall x.A}}{
    \vlin{\weakr}{}{\sqns{\Gamma, A}}{
      \vlhy{\sqns{\Gamma}}}}}
\leadsto
\vlderivation{
  \vlin{\weakr}{}{\sqns{\Gamma, \forall x.A}}{
    \vlhy{\sqns{\Gamma}}}}
\end{equation*}
\begin{equation*}
\vlderivation{
  \vlin{\conr}{}{\sqns{\Gamma, A}}{
    \vlin{\weakr}{}{\sqns{\Gamma, A, A}}{
      \vlhy{\sqns{\Gamma, A}}}}}
\leadsto
\vlderivation{
  \vlhy{\sqns{\Gamma, A}}}
\end{equation*}
Note that in the case of $\vlor$, we use the deep rule $\wrD$ which can be
permuted down over all the rules. By using these rewriting rules, we can eventually get
a derivation with all the instances of $\weakr$ and $\wrD$ at the bottom. Now observe
that the instances of $\conr$ in $\Deri$ can be transformed using the following
rule:

@@

\begin{proof}
  First, observe that the $\exists$ rule can be permuted downwards over all the
other rules since $A\ssubst{x}{t}$ has the same structure as $A$ and none of the
other rules has a premise of the form $\Scons{\exists x.A}$. It suffices now to
prove that for all $r_1 \in \set{\forall, \aiD, \tttD}$, for all $r_2 \in
\set{\sw, \mix, \fequ}$, we can permute $r_1$ upwards over $r_2$.
  We show some cases here, and leave the other to the reader.
  \begin{equation*}
  \vlderivation{
    \vlin{\aiD}{}{\Scons{(A \vlan (a \vlor \cneg a)) \vlor C}}{
      \vlin{\sw}{}{\Scons{(A \vlan \ttt) \vlor C}}{
        \vlhy{\Scons{A \vlan (\ttt \vlor C)}}}}}
  \leadsto
  \vlderivation{
    \vlin{\sw}{}{\Scons{(A \vlan (a \vlor \cneg a)) \vlor C}}{
      \vlin{\aiD}{}{\Scons{A \vlan ((a \vlor \cneg a) \vlor C)}}{
        \vlhy{\Scons{A \vlan (\ttt \vlor C)}}}}}
  \end{equation*}
  \begin{equation*}
    \vlderivation{
      \vlin{\tttD}{}{\Scons{(A \vlor (B \vlan \ttt))}}{
        \vlin{\mix}{}{\Scons{A \vlor B}}{
          \vlhy{\Scons{A \vlan B}}}}}
  \leadsto
  \vlderivation{
    \vlin{\mix}{}{\Scons{(A \vlor (B \vlan \ttt)}}{
      \vlin{\tttD}{}{\Scons{A \vlan (B \vlan \ttt)}}{
        \vlhy{\Scons{A \vlan B}}}}}
  \end{equation*}

@@

>>>>

  We have the following derivations:
  $$
  \vlderivation{
    \vlin{\cD}{}{\forall x.A}{
      \vlin{\equiv}{(x \notin fv(\forall x.A))}{(\forall x.A) \vlor (\forall x.A)}{
        \vlin{\wrD}{}{\forall x.((\forall x.A) \vlor A)}{
          \vlhy{\forall x.\forall x.A}}}}}
  $$.
  $$
  \vlderivation{
    \vlin{\cD}{}{(A \vlor B) \vlan (C \vlor D)}{
      \vlin{\fequ}{}{((A \vlor B) \vlan (C \vlor D)) \vlor ((A \vlor B) \vlan (C \vlor D))}{
      \vlin{\wrD}{}{((A \vlor B) \vlan (C \vlor D)) \vlor ((B \vlor A) \vlan (D \vlor C))}{
	\vlin{\wrD}{}{((A \vlor B) \vlan (C \vlor D)) \vlor ((B \vlor A) \vlan D))}{
	  \vlin{\wrD}{}{((A \vlor B) \vlan (C \vlor D)) \vlor (B \vlan D))}{
	    \vlin{\wrD}{}{((A \vlor B) \vlan C) \vlor (B \vlan D)}{
	      \vlhy{(A \vlan C) \vlor (B \vlan D)}}}}}}}}
  $$.
  $$
  \vlderivation{
    \vlin{\cD}{}{\exists x.(A \vlor B)}{
      \vlin{\fequ}{}{\exists x.(A \vlor B)) \vlor (\exists x.(A \vlor B))}{
      \vlin{\wrD}{}{(\exists x.(A \vlor B)) \vlor (\exists x.(B \vlor A))}{
	\vlin{\wrD}{}{(\exists x.(A \vlor B)) \vlor (\exists x.B)}{
	  \vlhy{(\exists x.A) \vlor (\exists x.B)}}}}}}
  $$.
  $$
  \vlderivation{
    \vlin{\cD}{}{\forall x.(A \vlor B)}{
      \vlin{\fequ}{}{(\forall x.(A \vlor B)) \vlor (\forall x.(A \vlor B))}{
      \vlin{\wrD}{}{(\forall x.(A \vlor B)) \vlor (\forall x.(B \vlor A))}{
	\vlin{\wrD}{}{(\forall x.(A \vlor B)) \vlor (\forall x.B)}{
	  \vlhy{(\forall x.A) \vlor (\forall x.B)}}}}}}
  $$.
  \juihsuan{done. If needed, we can introduce the notion of open deduction to
reduce the size of derivations...}\lutz{I was thinking about that, but (i) it is probably not worth the effort, as we wont have many derivation, and (ii) it is hard to define rectified derivations this way.}
\end{proof}

@@

@@

\begin{equation*}
\vlderivation{
  \vliin{\vlan}{}{\sqns{\Gamma, A \vlan B, \Delta}}{
    \vlin{\weakr}{}{\sqns{\Gamma, A}}{
      \vlhy{\sqns{\Gamma}}}}{
    \vlhy{\sqns{B, \Delta}}}}
\leadsto
\vlderivation{
  \vlin{\weakr}{}{\sqns{\Gamma, A \vlan B, \Delta}}{
    \vlin{\weakr}{}{\vlvdots}{
      \vlhy{\sqns{\Gamma}}}}}
\end{equation*}
\begin{equation*}
\vlderivation{
  \vlin{\vlor}{}{\sqns{\Gamma, A \vlor B}}{
    \vlin{\weakr}{}{\sqns{\Gamma, A, B}}{
      \vlhy{\sqns{\Gamma, A}}}}}
\leadsto
\vlderivation{
  \vlin{\wrD}{}{\sqns{\Gamma, A \vlor B}}{
    \vlhy{\sqns{\Gamma, A}}}}
\end{equation*}
\begin{equation*}
\vlderivation{
  \vlin{\exists}{}{\sqns{\Gamma, \exists x.A}}{
    \vlin{\weakr}{}{\sqns{\Gamma, A\ssubst{x}{t}}}{
      \vlhy{\sqns{\Gamma}}}}}
\leadsto
\vlderivation{
  \vlin{\weakr}{}{\sqns{\Gamma, \exists x.A}}{
    \vlhy{\sqns{\Gamma}}}}
\end{equation*}
\begin{equation*}
\vlderivation{
  \vlin{\forall}{}{\sqns{\Gamma, \forall x.A}}{
    \vlin{\weakr}{}{\sqns{\Gamma, A}}{
      \vlhy{\sqns{\Gamma}}}}}
\leadsto
\vlderivation{
  \vlin{\weakr}{}{\sqns{\Gamma, \forall x.A}}{
    \vlhy{\sqns{\Gamma}}}}
\end{equation*}
\begin{equation*}
\vlderivation{
  \vlin{\conr}{}{\sqns{\Gamma, A}}{
    \vlin{\weakr}{}{\sqns{\Gamma, A, A}}{
      \vlhy{\sqns{\Gamma, A}}}}}
\leadsto
\vlderivation{
  \vlhy{\sqns{\Gamma, A}}}
\end{equation*}
Note that in the case of $\vlor$, we use the deep rule $\wrD$ which can be
permuted down over all the rules. By using these rewriting rules, we can eventually get
a derivation with all the instances of $\weakr$ and $\wrD$ at the bottom. Now observe
that the instances of $\conr$ in $\Deri$ can be transformed using the following
rule:

@@

\begin{proof}
  First, observe that the $\exists$ rule can be permuted downwards over all the
other rules since $A\ssubst{x}{t}$ has the same structure as $A$ and none of the
other rules has a premise of the form $\Scons{\exists x.A}$. It suffices now to
prove that for all $r_1 \in \set{\forall, \aiD, \tttD}$, for all $r_2 \in
\set{\sw, \mix, \fequ}$, we can permute $r_1$ upwards over $r_2$.
  We show some cases here, and leave the others to the reader.
  \begin{equation*}
  \vlderivation{
    \vlin{\aiD}{}{\Scons{(A \vlan (a \vlor \cneg a)) \vlor C}}{
      \vlin{\sw}{}{\Scons{(A \vlan \ttt) \vlor C}}{
        \vlhy{\Scons{A \vlan (\ttt \vlor C)}}}}}
  \leadsto
  \vlderivation{
    \vlin{\sw}{}{\Scons{(A \vlan (a \vlor \cneg a)) \vlor C}}{
      \vlin{\aiD}{}{\Scons{A \vlan ((a \vlor \cneg a) \vlor C)}}{
        \vlhy{\Scons{A \vlan (\ttt \vlor C)}}}}}
  \end{equation*}
  \begin{equation*}
    \vlderivation{
      \vlin{\tttD}{}{\Scons{(A \vlor (B \vlan \ttt))}}{
        \vlin{\mix}{}{\Scons{A \vlor B}}{
          \vlhy{\Scons{A \vlan B}}}}}
  \leadsto
  \vlderivation{
    \vlin{\mix}{}{\Scons{(A \vlor (B \vlan \ttt)}}{
      \vlin{\tttD}{}{\Scons{A \vlan (B \vlan \ttt)}}{
        \vlhy{\Scons{A \vlan B}}}}}
  \end{equation*}

@@

\begin{proof}
  First, a derivation consisting of an instance of $\wrD$ followed by $\rr \in \set{\acD, \cfaD,
\me, \mfaD, \mexD}$ can be either replaced by a derivation consisting of $\wrD$
only or the instance of $\wrD$ can be permuted downwards. We show some cases
here, and leave the others to the reader.
  \begin{equation*}
  \vlderivation{
    \vlin{\mfaD}{}{\Scons{\forall x.(A \vlor B)}}{
      \vlin{\wrD}{}{\Scons{(\forall x.A) \vlor (\forall x.B)}}{
        \vlhy{\Scons{\forall x.A}}}}}
  \leadsto
  \vlderivation{
    \vlin{\wrD}{}{\Scons{\forall x.(A \vlor B)}}{
      \vlhy{\Scons{\forall x.A}}}}
  \end{equation*}
  \begin{equation*}
  \vlderivation{
    \vlin{\me}{}{\Scons{(A \vlor B) \vlan (C \vlor D)}}{
      \vlin{\wrD}{}{\Scons{(A \vlan C) \vlor (B \vlan D)}}{
        \vlhy{\Scons{A \vlan C}}}}}
  \leadsto
  \vlderivation{
    \vlin{\wrD}{}{\Scons{(A \vlor B) \vlan (C \vlor D)}}{
      \vlin{\wrD}{}{\Scons{(A \vlor B) \vlan C}}{
        \vlhy{\Scons{A \vlan C}}}}}
  \end{equation*}
  \begin{equation*}
  \vlderivation{
    \vlin{\acD}{}{\Scons{a}}{
      \vlin{\wrD}{}{\Scons{a \vlor a}}{
        \vlhy{\Scons{a}}}}}
  \leadsto
  \vlderivation{
    \vlhy{\Scons{a}}}
  \end{equation*}

  For $\rr_1 \in \set{\me, \mfaD, \mexD}$, $\rr_2 \in \set{\acD, \cfaD}$,
$\rr_1$ can be permuted upwards over $\rr_2$ (with some $\fequ$ inserted). The only non-trivial case is shown below:
  \begin{equation*}
  \vlderivation{
    \vlin{\mfaD}{}{\Scons{\forall x.(A \vlor B)}}{
      \vlin{\cfaD}{}{\Scons{(\forall x.A) \vlor (\forall x.B)}}{
        \vlhy{\Scons{(\forall x.\forall x.A) \vlor (\forall x.B)}}}}}
  \leadsto
  \vlderivation{
    \vlin{\mfaD}{}{\Scons{\forall x.(A \vlor B)}}{
      \vlin{\fequ}{}{\Scons{(\forall x.A) \vlor (\forall x.B)}}{
        \vlin{\mfaD}{}{\Scons{\forall x.(\forall x.A \vlor B)}}{
          \vlhy{\Scons{(\forall x.\forall x.A) \vlor (\forall x.B)}}}}}}
  \end{equation*}
  \todo{permutation with $\fequ$}
\end{proof}
